Fail Faster

### 2025-03-20 paper/paper.tex
interleave

\subsection{Benchmarking Generator Speed}
\label{subsection:benchmarking}
To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary
the choice of \rand in our \name generators, using both \bq's default
\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
in \secref{subsection:faster-rng}.

Our benchmarks consist of boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a custom, correct-by-construction generator of
STLC terms, whose behavior is the most complex.

The runtime of these generators over a range of sizes is shown in
\figref{fig:time}. Staging provides only a negligible speedup for type-derived
and insertion-based BST strategies, whose overhead is dominated by random
sampling. In this case, performance improvement is primarily attributable the
\csplitmix \rand.  In contrast, STLC strategies benefit more from staging, as
they involve fewer sampling operations but rely heavily on generator
combinators. The same holds for boolean lists, as sampling booleans is
inexpensive relative to integers. Our single-pass BST strategy falls somewhere
in between: it both samples frequently and makes extensive use of generator
combinators. \tr{Goal is to replace this paragraph with the ref-counting
chart---I think it will be WAY less boring. But I need to do 3.7 first.}

@@

\subsubsection{Scala} To show that the benefit of staging extends to
other strict functional languages, we apply it to ScalaCheck, a widely-used PBT
library written in Scala. We implement a subset of our strategies from
\secref{subsection:benchmarking} in both standard and staged ScalaCheck
(``\scalaname'').

As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain
over ScalaCheck than \name does over \bq. This discrepancy arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging allows us to eliminate this
overhead, resulting in substantial speedups in aggregate.

These results illustrate that staging is a powerful \textit{and} general optimization technique, capable of delivering performance gains in strict functional languages beyond OCaml.
\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. Performance benefits scale with generation size; lower is better. BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix.}
  \label{fig:time}
\end{figure}

>>>>

\subsection{Benchmarking Generator Speed}
\label{subsection:benchmarking}
To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to identical ones in \bq,
and generators in \scalaname to those in ScalaCheck.
 We vary the choice of \rand in our \camlname generators, using both \bq's default
\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
in \secref{subsection:faster-rng}.

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a custom, correct-by-construction generator of
STLC terms, whose behavior is the most complex. 
For evaluating \scalaname, we implement a subset of the afformentioned strategies  ---
only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.

The runtime of the \camlname generators over a range of sizes is shown in
\figref{fig:time}. Staging provides only a negligible speedup for type-derived
and insertion-based BST strategies, whose overhead is dominated by random
sampling. In this case, performance improvement is primarily attributable the
\csplitmix \rand.  In contrast, STLC strategies benefit more from staging, as
they involve fewer sampling operations but rely heavily on generator
combinators. The same holds for boolean lists, as sampling booleans is
inexpensive relative to integers. Our single-pass BST strategy falls somewhere
in between: it both samples frequently and makes extensive use of generator
combinators. \tr{Goal is to replace this paragraph with the ref-counting
chart---I think it will be WAY less boring. But I need to do 3.7 first.}

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain
--- purely from staging --- over ScalaCheck than \name does over \bq. This discrepancy arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging allows us to eliminate this
overhead, resulting in substantial speedups in aggregate.

@@

% \subsubsection{Scala} To show that the benefit of staging extends to
% other strict functional languages, we apply it to ScalaCheck, a widely-used PBT
% library written in Scala.

\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. Performance benefits scale with generation size; lower is better. BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix.}
  \label{fig:time}
\end{figure}
